A data-driven, pixel-level model to improve the precision of \Kepler\ photometry

### 2013-10-22 documents/plm.tex
issue for dfm
closing DFM-related issues

In the context of this project, the quantities of interest are the light-curves
  (photometry as a function of time or apparent flux as a function of time)
  for every star.
What we get, for every star, from the data-driven model, is a train-and-test prediction
  for the brightness of every pixel in the patch of the detector relevant to the star,
  along with the residual, or the difference between the measured read-out pixel value and the prediction (data minus prediction).
What will be called the ``\name\ photometry'' in what follows is constructed from what will be called the ``\name\ images''.
The \name\ photometry is constructed from the \name\ images
  just as the \Kepler\ official SAP photometry is constructed from the \todo{XXX images};
  that is, by summing pre-defined sets of pixels (with unit weights; see above and below for criticism of this).
The \name\ images are constructed from the data-driven model predictions by taking,
  for each pixel in each contiguous data segment (see below),
  the \emph{mean} predicted pixel value (mean over time),
  and adding to it the residuals (data minus prediction) away from the data-driven model prediction.
These \name\ images are ``calibrated'' images,
  in that the data-driven model has been used to fit out all the variability it can,
  subject to the restrictions imposed by its finite freedom and the train-and-test specification.
The signal---in which exoplanet transits will be found---is fundamentally the observed residual away from this prediction,
  but it has to be added back into the mean prediction to have the right ``baseline'' or zero level.
In short, \name\ will produce prediction images, residual images, \name\ images (mean prediction plus residual),
  and aperture-based \name\ photometry analogous to the SAP photometry.

Finally, the fifth general idea is that there is no objective \emph{ground truth} against which we can tell
  that any particular data-driven model is better or worse than any other.
This problem is a problem for \name, and for the \Kepler\ PDC, and any other data-driven calibration models.
One might think that the ``best'' model is the one that predicts data with lowest variance;
  this would be true if all models adhered to the same train-and-test framework, which they don't.
Besides, a model that can predict not just the spacecraft-induced variability,
  but also variability caused by exoplanet transits of interest,
  will effectively over-fit and distort the most important information in the light-curves.
That is, what is considered best for modeling the data depends on the objectives of the user.
For us, who are interested (in the long term) in finding and characterizing Earth analogs,
  the ``best'' data-driven model is the one that produces the most success in finding and characterizing Earth analogs!
What we will show in what follows is that the \name\ photometry does not distort
  artificial exoplanet transit signals injected into real \Kepler\ pixel-level data.
We can also show that it produces stellar light-curves that look sensible for known variable stars.
We cannot yet show that it is optimal for our long-term goals,
  nor can we show that it produces images or light-curves that are more accurately calibrated;
  we will only show promising properties of the \name\ outputs.
In the end, the value of the \name\ will be demonstrated by the scientific projects it enables.

@@

Let's consider a particular pixel $m$ in the focal plane in one particular quarter.
In that quarter, the spacecraft delivers $N\sim 4000$ measurements $I_{mn}$
  of the intensity falling in that pixel at the $N$ times $t_n$ at which exposures were taken during the quarter.
(For intensity measurements $I_{mn}$ here we are using pixels from the \todo{WHICH \Kepler\ image product?})
We want to build a prediction for the intensity $I_{mn}$ in pixel $m$ at each time $t_n$
  using the intensities $I_{m'n}$ of other pixels $m'$.
The question is:  What other pixels to choose?
There are many possible qualitatively and quantitatively different choices here.
Somewhat arbitrarily,
  we choose all the pixels $m'$ associated with the $Q=5$ stars on the same CCD in the same quarter
  that are closest in magnitude
  (\Kepler\ magnitude as reported in the \Kepler\ Input Catalog)
  to the target star associated with pixel $m$.
This set of pixels%
  ---from the $Q$ stars on the same CCD---%
  is the \emph{pixel set} $\set{M}_m$ associated with pixel $m$.
Note that because the set $\set{M}_m$ is of pixels associated with \emph{different} stars
  than the star associated with pixel $m$,
  the pixel $m$ will not be in the set $\set{M}_m$,
  and nor will any of pixel $m$'s close neighbors.
That is, there will be no (or almost no) overlap in stellar illumination of pixel $m$
  and the pixels $m'\in\set{M}_m$.

@@

As mentioned above, in the context of this project,
  a model is something that can predict data given settings of some parameters,
  and an objective function that can be used to find best values for those parameters.
We prefer to work in a probabilistic mode in which the objective function can be justified
  in terms of a likelihood (probability for the data given the model)
  or a posterior probability density function (likelihood times a prior pdf).
The PLM treats each \Kepler\ data point $I_{mn}$ as being
  predictable from a linear combination of data points $I_{m'n}$,
  where $m'$ is from the set of non-overlapping pixels $m'\in\set{M}_m$.
\begin{eqnarray}
I_{mn}         &=& I^{\ast}_{mn} + e_{mn}
\\
I^{\ast}_{mn}  &=& \sum_{m'\in\set{M}_m} a_{mm'}\,I_{m'n} + \sum_{k=1}^{K} b_{mk}\,d_{kn}
\quad,
\end{eqnarray}
where $I^{\ast}_{mn}$ is the prediction (by the model) for data point $I_{mn}$,
  $e_{mn}$ is a noise contribution or residual away from the prediction,
  and the $a_{mm'}$ and $b_{mk}$ are the parameters (linear coefficients of the prediction).
If we presume that the residuals away from the prediction are normally distributed with zero mean
  and known variance,
  likelihood optimization reduces to chi-squared minimization.
We add to the standard chi-squared definition a regularization term
  (equivalent to multiplying the likelihood by a Gaussian prior pdf)
  that penalizes large absolute values for the the coefficients $a_{mm'}$:
\begin{eqnarray}
\chi^2_{mn}    &=& \sum_{n'\in\set{N}_n} \frac{[I_{mn'} - I^{\ast}_{mn'}]^2}{\sigma^2_{mn'}}
                 + \sum_{m'\in\set{M}_m} \frac{a_{mm'}^2}{\Sigma^2}
\\
I^{\ast}_{mn'} &=& \sum_{m'\in\set{M}_m} a_{mm'}\,I_{m'n'} + \sum_{k=1}^{K} b_{mk}\,d_{kn'}
\quad,
\end{eqnarray}
where the $\sigma^2_{mn'}$ are the (presumed known) individual-pixel noise variances,
  and $\Sigma^2$ sets the strength of the regularization (or width of the prior pdf).
\todo{DFM: We set or ``know'' the variances how?
  I think we should use the inverse of the median of the reported inverse variances.
  Also, we set the prior strength to what value?
  Heuristic justification of the latter.}

The train-and-test idea comes into the objective function $\chi^2_{mn}$:
When we are computing the objective function $\chi^2_{mn}$,
  we are using only time points $n'\in\set{N}_n$ that don't overlap the target (or test) time point $n$.
We train the model---that is, set the parameters $a_{mm'}$ and $b_{mk}$---%
  by choosing the full set of parameters that jointly minimize the objective function $\chi^2_{mn}$.
Importantly, and perhaps surprisingly, this is a \emph{different objective function}
  for every target data point (pixel value to be predicted) $I_{mn}$,
  and we have to do a \emph{independent optimization} of the parameters for every pixel value we want to predict.
That is, every pixel datum $I_{mn}$ we predict will have
  \emph{different} settings of the parameters $a_{mm'}$ and $b_{mk}$.
Since there are about $10^{11}$ total pixel values in the \Kepler\ \todo{what imaging?} data archive,
  this represents one fuck of a lot of optimizations,
  and a lot of parameters to obtain and record.

@@

\todo{HOGG: We construct the predicted images and the \name\ images how?}

\todo{HOGG: We construct the \name\ photometry from the \name\ images how?}

@@

\clearpage
\begin{figure}
~ % DFM: replace this ~ with a figure
\caption{A patch of \Kepler\ imaging from \todo{XXX quarter}, centered on \todo{YYY target}.
  In the top panel, each box represents a pixel.
  Within each box is shown the light-curve of that pixel over the quarter.
  In the bottom panel, the SAP photometry for this target is shown;
  the SAP photometry is based on a straight sum of the pixels marked with darker borders.
  Note that every pixel varies substantially but the total SAP photometry is better than any individual pixel.
  Note also that the pixel variations are strongly inter-related.\label{fig:pixelpatch}}
\end{figure}

>>>>

In the context of this project, the quantities of interest are the light-curves
  (photometry as a function of time or apparent flux as a function of time)
  for every star.
What we get, for every star, from the data-driven model, is a train-and-test prediction
  for the brightness of every pixel in the patch of the detector relevant to the star,
  along with the residual, or the difference between the measured read-out pixel value and the prediction (data minus prediction).
What will be called the ``\name\ photometry'' in what follows is constructed from what will be called the ``\name\ pixels''.
The \name\ photometry is constructed from the \name\ pixels
  just as the \Kepler\ official SAP photometry is constructed from the \Kepler\ Target Pixel Files;
  that is, by summing pre-defined sets of pixels (with unit weights; see above and below for criticism of this).
The \name\ pixels are constructed from the data-driven model predictions by taking,
  for each pixel in each contiguous data segment (see below),
  the \emph{mean} predicted pixel value (mean over time),
  and adding to it the residuals (data minus prediction) away from the data-driven model prediction.
These \name\ pixels are ``calibrated'' pixels,
  in that the data-driven model has been used to fit out all the variability it can,
  subject to the restrictions imposed by its finite freedom and the train-and-test specification.
The signal---in which exoplanet transits will be found---is fundamentally the observed residual away from this prediction,
  but it has to be added back into the mean prediction to have the right ``baseline'' or zero level.
In short, \name\ will produce prediction pixels, residual pixels, \name\ pixels (mean prediction plus residual),
  and aperture-based \name\ photometry analogous to the SAP photometry.

Finally, the fifth general idea is that there is no objective \emph{ground truth} against which we can tell
  that any particular data-driven model is better or worse than any other.
This problem is a problem for \name, and for the \Kepler\ PDC, and any other data-driven calibration models.
One might think that the ``best'' model is the one that predicts data with lowest variance;
  this would be true if all models adhered to the same train-and-test framework, which they don't.
Besides, a model that can predict not just the spacecraft-induced variability,
  but also variability caused by exoplanet transits of interest,
  will effectively over-fit and distort the most important information in the light-curves.
That is, what is considered best for modeling the data depends on the objectives of the user.
For us, who are interested (in the long term) in finding and characterizing Earth analogs,
  the ``best'' data-driven model is the one that produces the most success in finding and characterizing Earth analogs!
What we will show in what follows is that the \name\ photometry does not distort
  artificial exoplanet transit signals injected into real \Kepler\ pixel-level data.
We can also show that it produces stellar light-curves that look sensible for known variable stars.
We cannot yet show that it is optimal for our long-term goals,
  nor can we show that it produces pixelss or light-curves that are more accurately calibrated;
  we will only show promising properties of the \name\ outputs.
In the end, the value of the \name\ will be demonstrated by the scientific projects it enables.

@@

Let's consider a particular pixel $m$ in the focal plane in one particular quarter.
In that quarter, the spacecraft delivers $N\sim 4000$ measurements $I_{mn}$
  of the intensity falling in that pixel at the $N$ times $t_n$ at which exposures were taken during the quarter.
(For intensity measurements $I_{mn}$ here we are using pixels from the Target Pixel Files.)
We want to build a prediction for the intensity $I_{mn}$ in pixel $m$ at each time $t_n$
  using the intensities $I_{m'n}$ of other pixels $m'$.
The question is:  What other pixels to choose?
There are many possible qualitatively and quantitatively different choices here.
Somewhat arbitrarily,
  we choose all the pixels $m'$ associated with the $Q=5$ stars on the same CCD in the same quarter
  that are closest in magnitude
  (\Kepler\ magnitude as reported in the \Kepler\ Input Catalog)
  to the target star associated with pixel $m$.
This set of pixels%
  ---from the $Q$ stars on the same CCD---%
  is the \emph{pixel set} $\set{M}_m$ associated with pixel $m$.
Note that because the set $\set{M}_m$ is of pixels associated with \emph{different} stars
  than the star associated with pixel $m$,
  the pixel $m$ will not be in the set $\set{M}_m$,
  and nor will any of pixel $m$'s close neighbors.
That is, there will be no (or almost no) overlap in stellar illumination of pixel $m$
  and the pixels $m'\in\set{M}_m$.

@@

As mentioned above, in the context of this project,
  a model is something that can predict data given settings of some parameters,
  and an objective function that can be used to find best values for those parameters.
We prefer to work in a probabilistic mode in which the objective function can be justified
  in terms of a likelihood (probability for the data given the model)
  or a posterior probability density function (likelihood times a prior pdf).
The PLM treats each \Kepler\ data point $I_{mn}$ as being
  predictable from a linear combination of data points $I_{m'n}$,
  where $m'$ is from the set of non-overlapping pixels $m'\in\set{M}_m$.
\begin{eqnarray}
I_{mn}         &=& I^{\ast}_{mn} + e_{mn}
\\
I^{\ast}_{mn}  &=& \sum_{m'\in\set{M}_m} a_{mnm'}\,I_{m'n} + \sum_{k=1}^{K} b_{mnk}\,d_{kn}
\quad,
\end{eqnarray}
where $I^{\ast}_{mn}$ is the prediction (by the model) for data point $I_{mn}$,
  $e_{mn}$ is a noise contribution or residual away from the prediction,
  and the $a_{mnm'}$ and $b_{mnk}$ are the parameters (linear coefficients of the prediction).
The parameters $a_{mnm'}$ and $b_{mnk}$ have indices $m$ because
  they are different for every pixel $m$,
  and they will even be different for every time step $n$;
  that is, there will be a separate best-fit value for the parameters for every $I_{mn}$.
If we presume that the residuals away from the prediction are normally distributed with zero mean
  and known variance,
  likelihood optimization reduces to chi-squared minimization.
We add to the standard chi-squared definition a regularization term
  (equivalent to multiplying the likelihood by a Gaussian prior pdf)
  that penalizes large absolute values for the the coefficients $a_{mnm'}$:
\begin{eqnarray}
\chi^2_{mn}    &=& \sum_{n'\in\set{N}_n} \frac{[I_{mn'} - I^{\ast}_{mnn'}]^2}{\sigma^2_{mn'}}
                 + \sum_{m'\in\set{M}_m} \frac{a_{mnm'}^2}{\Sigma^2}
\\
I^{\ast}_{mnn'} &=& \sum_{m'\in\set{M}_m} a_{mnm'}\,I_{m'n'} + \sum_{k=1}^{K} b_{mnk}\,d_{kn'}
\quad,
\end{eqnarray}
where the $\sigma^2_{mn'}$ are the (presumed known) individual-pixel noise variances,
  and $\Sigma^2$ sets the strength of the regularization (or width of the prior pdf).
Although the variances $\sigma^2_{mn'}$ are in principle different for every time step $n'$,
  for nearly-constant stars in good data, there is not much variation;
  in what follows, we treat these uncertainties as fixed at
  \todo{DFM: some mean value (we can't ignore this,
   because it ``plays off against'' the regularization)}
The regularization variance $\Sigma^2$ we set to 0.3 (\todo{DFM TAKE NOTE})
  because we expect almost all the coefficients $a_{mnm'}$ to be less than unity in magnitude.
We do not regularize the meta-data coefficients $b_{mnk}$,
  because we want the model to make use of the (easily interpretable) meta-data coefficients as much as possible.

The train-and-test idea comes into the objective function $\chi^2_{mn}$:
When we are computing the objective function $\chi^2_{mn}$,
  we are using only time points $n'\in\set{N}_n$ that don't overlap the target (or test) time point $n$.
We train the model---that is, set the parameters $a_{mnm'}$ and $b_{mnk}$---%
  by choosing the full set of parameters that jointly minimize the objective function $\chi^2_{mn}$.
Importantly, and perhaps surprisingly, this is a \emph{different objective function}
  for every target data point (pixel value to be predicted) $I_{mn}$,
  and we have to do a \emph{independent optimization} of the parameters for every pixel value we want to predict.
That is, every pixel datum $I_{mn}$ we predict will have
  \emph{different} settings of the parameters $a_{mnm'}$ and $b_{mnk}$.
Since there are about $10^{11}$ total pixel values in the \Kepler\ Target Pixel Files,
  this represents one fuck of a lot of optimizations,
  and a lot of parameters to obtain and record.

@@

Once we have obtained the $a_{mnm'}$ parameters corresponding to data point $I_{mn}$ by minimizing the training objective $\chi^2_{mn}$.

\todo{HOGG: We construct the predicted pixels and the \name\ pixels how?}

\todo{HOGG: We construct the \name\ photometry from the \name\ pixels how?}

@@

\clearpage
\begin{figure}
~ % DFM: replace this ~ with a figure
\caption{A patch of \Kepler\ pixels from the Target Pixel Files
  from \todo{XXX quarter}, centered on \todo{YYY target}.
  In the top panel, each box represents a pixel.
  Within each box is shown the light-curve of that pixel over the quarter.
  In the bottom panel, the SAP photometry for this target is shown;
  the SAP photometry is based on a straight sum of the pixels marked with darker borders.
  Note that every pixel varies substantially but the total SAP photometry is better than any individual pixel.
  Note also that the pixel variations are strongly inter-related.\label{fig:pixelpatch}}
\end{figure}
